Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values

### 2020-11-01 doc_arXiv/doc_arXiv.tex
all tables and figs added

The Interagency Registry for Mechanically Assisted Circulatory Support
(INTERMACS) was launched to improve MCS patient outcomes through the
collection of patient characteristics, medical events, and long terms
outcomes at a nation wide level. Currently, INTERMACS comprises over
20,000 patients who have received a MCS device. As the largest registry
for data on patients receiving MCS devices, INTERMACS has been leveraged
to develop numerous risk prediction models for mortality and other types
of adverse events that may occur after receiving a device
\cite{kirklin2017eighth, kormos2019society, Adamo950}. A critical step
for developing risk prediction models is handling missing data, which is
common in a ``real world'' database that is not designed for the rigor
of data completeness found in a clinical trial. While some missing
values are related to data entry error, others may relate to instability
of a patient's circulatory status at the time of implant. For instance,
hemodynamic laboratory values are missing for patients in whom invasive
catheterizations were not performed. Patients without invasive
cathaderizations may be too ill (unstable) to tolerate the procedure.
Instead, such patients may, after determination of basic hemodynamics,
proceed directly to device placement.

@@

INTERMACS is a North American observational registry for patients
receiving MCS devices that began as a partnership between the National
Heart, Lung, and Blood Institute, US Food and Drug Administration, the
Centers for Medicaid and Medicare Services, industry, and individual
hospitals with the mission of improving MCS outcomes. In 2018, INTERMACS
became an official Society of Thoracic Surgeons database.

@@

Though the outcome variable may be useful for imputing predictor values,
using outcome variables for this purpose in supervised learning projects
can lead to issues in clinical implementation. For instance, suppose we
seek to predict an outcome \(Y\) for a patient in the clinical setting.
If the patient is missing information for a predictor \(X\) and our only
method to impute missing values of \(X\) leverages the observed value of
\(Y\), how should we impute \(X\)? If we do not already know \(Y\),
which would seem likely for clinical application of a prediction model,
then we cannot impute \(X\). On the other hand, if we already know
\(Y\), then we obviously do not need to predict it. Due to these
practical considerations and because INTERMACS is often leveraged to
create prediction models for clinical practice \cite{thomas2014pre}, we
did not leverage outcome variables (\textit{i.e., }time until
transplant, death, or last contact) to impute missing values of
predictors in the current analysis.

@@

KNN imputation identifies \(k\) `similar' observations
(\textit{i.e., }neighbors) for each observation with a missing value in
a given variable \cite{chen2000nearest}. A donor value for the current
missing value is generated by sampling or aggregating the values from
the \(k\) nearest neighbors. In the current analysis, we identified 10
nearest neighbors using Gower's distance \cite{gower}. When imputing a
single dataset using KNN, we aggregated values from nearest neighbors
using the median for numeric variables and the mode for categorical
variables. When imputing multiple datasets using KNN, we sampled one
value at random from nearest neighbors for each imputed set.

@@

The Brier score is dependent on the rate of observed events, which can
make it a difficult metric to interpret. It is often more informative to
scale the Brier score based on the Brier score of a naive model. More
specifically, for a given risk prediction model, the scaled Brier score
is computed as
\[\textrm{Scaled } \widehat{\textrm{BS}}(t)\textrm{ of model } = 1 - \frac{\widehat{\textrm{BS}}(t)\textrm{ of model}}{\widehat{\textrm{BS}}(t)\textrm{ of naive model}}.\]
As the Brier score for risk prediction is analogous to mean-squared
error for prediction of a continuous outcome, the scaled Brier score is
analogous to the \(R^2\) statistic. Similar to the \(R^2\) statistic, a
scaled \(\widehat{\textrm{BS}}(t)\) of 1.00 and 0.00 indicate a perfect
and worthless model, respectively. In our analyses, a Kaplan-Meier
estimate based on the training data (\textit{i.e., }a risk prediction
model that did not use any predictor variables) provided the naive
prediction.

@@

To assess the prognostic value of each missing data strategy, we
internally validated a total of FILL IN modeling algorithm based on
combinations of imputation strategies and modeling strategies described
in Sections \ref{subsec:imputation} and \ref{subsec:modeling}. For
convenience, we use the term `modeling algorithm' to denote the
combination of a missing data strategy and a modeling strategy
(\textit{e.g., }imputation using mean/mode values followed by fitting
the PH model with stepwise variable selection) \cite{kuhn2013applied}.
We conducted internal validation using 200 replicates of Monte-Carlo
cross validation (MCCV), a resampling technique for internal validation
(Figure \ref{fig:monte_carlo_cv}).

\paragraph{Steps taken in each MCCV replicate}

In each replicate, 50\% of the available data were used for model
training and testing. Imputation was conducted in the training and
testing sets, separately, for each imputation strategy. Some imputation
strategies (\textit{e.g., }KNN and random forests) used only the
training data to generate imputed values, whereas others
(\textit{e.g., }Bayesian regression, PMM, and hot deck)

After imputation, each modeling algorithm (\textit{i.e., }stepwise PH
models, GBDTs, and the cIF) was applied, separately, to develop a RPE
for death or transplant that incorporated up to 50 predictor variables
from the entire collection of covariates available.

@@

Participant characteristics were tabulated for the overall population
and stratified by event status. Continuous and categorical variables
were summarized as mean (standard deviation) and percent, respectively.
Imputation accuracy was aggregated for all numeric and nominal variables
to create two overall scores for each imputed dataset, and the
distribution of these scores was tabulated for each imputation strategy.

@@

implement \texttt{miceRanger}, an R package similar to
\texttt{missForest} that implements MICE with random forests by
implementing PMM \cite{missForest}. \texttt{miceRanger} follows an
iterative procedure where one random forest is fit to each variable in a
dataset, separately, and then used to impute missing values in that
variable. The procedure may repeat until estimated generalization error
of the random forest collection stabilizes.

@@

@@

@@

\begin{tabular}{lccccc}
\toprule
Characteristic & Overall & Censored & Cessation of support & Dead & Transplant\\
\midrule
No. of patients & 14,738 & 6,308 & 310 & 4,204 & 3,916\\
Follow up time, months & 18 (16) & 25 (18) & 16 (11) & 13 (14) & 13 (10)\\
Age, years & 57 (13) & 57 (13) & 48 (14) & 61 (12) & 53 (12)\\
Sex &  &  &  &  & \\
\hspace{1em}Female & 21 & 22 & 41 & 22 & 19\\
\addlinespace
\hspace{1em}Male & 79 & 78 & 59 & 78 & 81\\
Race &  &  &  &  & \\
\hspace{1em}White & 66 & 63 & 64 & 71 & 67\\
\hspace{1em}Black & 25 & 28 & 27 & 21 & 23\\
\hspace{1em}Other & 9.2 & 9.6 & 9.4 & 7.4 & 10\\
\addlinespace
Body mass index & 29 (7) & 29 (7) & 29 (8) & 29 (7) & 28 (6)\\
Device strategy &  &  &  &  & \\
\hspace{1em}Other & 0.6 & 0.6 & 2.6 & 0.5 & 0.6\\
\hspace{1em}Bridge to transplant & 52 & 43 & 51 & 40 & 81\\
\hspace{1em}Destination therapy & 47 & 56 & 46 & 60 & 18\\
\addlinespace
Device type &  &  &  &  & \\
\hspace{1em}Bi-ventricular assistance device & 3.5 & 2.1 & 0.3 & 5.8 & 3.4\\
\hspace{1em}Left-ventricular assistance device & 97 & 98 & 100 & 94 & 97\\
\bottomrule
\multicolumn{6}{l}{\rule{0pt}{1em}\textit{Note: }}\\
\multicolumn{6}{l}{\rule{0pt}{1em}Table values are mean (standard deviation) or \%}\\
\end{tabular}

@@

@@

\begin{tabular}{lcccc}
\toprule
& \multicolumn{2}{c}{Nominal variables} & \multicolumn{2}{c}{Numeric variables} \\ 
 \cmidrule(lr){2-3}\cmidrule(lr){4-5}
 & Single imputation & Multiple imputation & Single imputation & Multiple imputation \\ 
\midrule
\multicolumn{1}{l}{+15\% additional missing data} \\ 
\midrule
Imputation to the mean & 0 (reference) & --- & 0 (reference) & --- \\ 
Hot deck & -0.24 (-0.34, -0.16) & -0.25 (-0.37, -0.18) & -0.35 (-0.53, -0.29) & -0.42 (-0.54, -0.36) \\ 
K-nearest-neighbors & 0.02 (-0.05, 0.15) & -0.19 (-0.29, -0.08) & 0.01 (-0.01, 0.03) & -0.33 (-0.42, -0.28) \\ 
Predictive mean matching & -0.15 (-0.34, 0.00) & -0.15 (-0.35, 0.00) & -0.25 (-0.40, -0.20) & -0.26 (-0.35, -0.21) \\ 
Random forest & -0.15 (-0.26, -0.02) & -0.15 (-0.26, -0.03) & 0.06 (0.02, 0.08) & 0.05 (0.02, 0.07) \\ 
Bayesian regression & -0.14 (-0.26, 0.01) & -0.14 (-0.27, 0.01) & -0.30 (-0.41, -0.23) & -0.30 (-0.41, -0.23) \\ 
\midrule
\multicolumn{1}{l}{+30\% additional missing data} \\ 
\midrule
Imputation to the mean & 0 (reference) & --- & 0 (reference) & --- \\ 
Hot deck & -0.25 (-0.34, -0.18) & -0.26 (-0.36, -0.18) & -0.34 (-0.46, -0.30) & -0.40 (-0.53, -0.36) \\ 
K-nearest-neighbors & -0.04 (-0.09, 0.03) & -0.24 (-0.34, -0.16) & -0.02 (-0.03, -0.01) & -0.36 (-0.44, -0.32) \\ 
Predictive mean matching & -0.18 (-0.38, -0.04) & -0.18 (-0.37, -0.04) & -0.29 (-0.46, -0.24) & -0.29 (-0.40, -0.25) \\ 
Random forest & -0.18 (-0.27, -0.07) & -0.18 (-0.28, -0.07) & -0.02 (-0.05, 0.01) & -0.01 (-0.04, 0.01) \\ 
Bayesian regression & -0.17 (-0.29, -0.04) & -0.17 (-0.29, -0.05) & -0.31 (-0.43, -0.27) & -0.31 (-0.43, -0.27) \\ 
\bottomrule
\end{tabular} 
 \end{table}

>>>>

The Interagency Registry for Mechanically Assisted Circulatory Support
(INTERMACS) was launched to improve MCS patient outcomes through the
collection of patient characteristics, medical events, and long terms
outcomes at a nation wide level. Currently, INTERMACS comprises over
20,000 patients who have received a MCS device. As the largest registry
for data on patients receiving MCS devices, INTERMACS has been leveraged
to develop numerous risk prediction models for mortality and other types
of adverse events that may occur after receiving a device
\cite{kirklin2017eighth, kormos2019society, Adamo950}. A critical step
for developing risk prediction models is handling missing data, which is
common in a ``real world'' database that is not designed for the rigor
of data completeness found in a clinical trial. While some missing
values are related to data entry error, others may relate to instability
of a patient's circulatory status at the time of implant. For instance,
hemodynamic laboratory values are missing for patients in whom invasive
catheterizations were not performed. Patients without invasive
catheterizations may be too ill (unstable) to tolerate the procedure.
Instead, such patients may, after determination of basic hemodynamics,
proceed directly to device placement.

@@

The INTERMACS data is publicly available on biolincc at
\url{https://biolincc.nhlbi.nih.gov/studies/intermacs/}. INTERMACS is a
North American observational registry for patients receiving MCS devices
that began as a partnership between the National Heart, Lung, and Blood
Institute, US Food and Drug Administration, the Centers for Medicaid and
Medicare Services, industry, and individual hospitals with the mission
of improving MCS outcomes. In 2018, INTERMACS became an official Society
of Thoracic Surgeons database.

@@

\paragraph{Using outcomes during imputation}

The outcome variable should be used to impute predictor values for
statistical inference, but using outcome variables for this purpose in
supervised learning can have unintended consequences in model
implementation. For instance, suppose we seek to predict an outcome
\(Y\) for a patient in the clinical setting. If the patient is missing
information for a predictor \(X\) and our model's strategy to impute
missing values of \(X\) leverages the observed value of \(Y\), how
should we impute \(X\)? If we do not already know \(Y\), then we cannot
impute \(X\) and hence cannot generate a valid prediction. On the other
hand, if we already know \(Y\), then we do not need to predict it. Due
to these practical considerations and because INTERMACS is often
leveraged to create prediction models for clinical practice
\cite{thomas2014pre}, we did not leverage outcome variables to impute
missing values of predictors in the current analysis.

@@

KNN imputation identifies \(k\) `similar' observations
(\textit{i.e., }neighbors) for each observation with a missing value in
a given variable \cite{chen2000nearest}. A donor value for the current
missing value is generated by sampling or aggregating the values from
the \(k\) nearest neighbors. In the current analysis, we identified 10
nearest neighbors using Gower's distance \cite{gower}. When imputing a
single dataset using KNN, we aggregated values from 10 nearest neighbors
using the median for numeric variables and the mode for categorical
variables. When imputing multiple datasets using KNN, we aggregated
values from 2, 6, 11, 16, and 20 neighbors, separately, to create 5
imputed datasets.

@@

The Brier score is dependent on the rate of observed events, which can
make it a difficult metric to interpret. It is often more informative to
scale the Brier score based on the Brier score of a naive model. More
specifically, for a given risk prediction model, the scaled Brier score
is computed as
\[\textrm{Scaled } \widehat{\textrm{BS}}(t)\textrm{ of model } = 1 - \frac{\widehat{\textrm{BS}}(t)\textrm{ of model}}{\widehat{\textrm{BS}}(t)\textrm{ of naive model}}.\]
As the Brier score for risk prediction is analogous to mean-squared
error for prediction of a continuous outcome, the scaled Brier score is
analogous to the \(R^2\) statistic. Similar to the \(R^2\) statistic, a
scaled \(\widehat{\textrm{BS}}(t)\) of 1.00 and 0.00 indicate a perfect
and worthless model, respectively. In our analyses, a Kaplan-Meier
estimate based on the training data (\textit{i.e., }a risk prediction
model that did not use any predictor variables) provided the naive
prediction. In the current analysis, we multiply scaled Brier score
values by 100 for ease of interpretation.

\paragraph{Discrimination and calibration}

The discrimination of a risk prediction model measures the probability
that the model will successfully identify which of two observations is
at higher risk for the event of interest. We estimated discrimination
using a time-dependent concordance (C-) index that accounted for
covariate-dependent censoring. A C-index of 0.50 and 1.00 correspond to
worthless and perfect discrimination, respectively. Similar to the
scaled Brier score, we multiply C-index values by 100 for ease of
interpretation. Calibration slope plots measure a risk prediction
model's absolute accuracy. We estimated calibration error by averaging
the squared distance between expected and observed event rates according
to a calibration plot. Full description of these evaluation metrics are
available \cite{gerds2014calibration, gerds2013estimating}.

@@

To assess the prognostic value of each missing data strategy, we
internally validated a total of 23 modeling algorithm based on
combinations of imputation strategies and modeling strategies described
in Sections \ref{subsec:imputation} and \ref{subsec:modeling}. For
convenience, we use the term `modeling algorithm' to denote the
combination of a missing data strategy and a modeling strategy
(\textit{e.g., }imputation using mean/mode values followed by fitting
the PH model with stepwise variable selection) \cite{kuhn2013applied}.
We conducted internal validation using 200 replicates of Monte-Carlo
cross validation, a resampling technique for internal validation.

\paragraph{Steps taken in each replicate}

In each replicate of Monte-Carlo cross validation, 50\% of the available
data were used for model training and testing. All predictor variables
with \textless{} 50\% missing values were considered for imputation and
subsequent model development. Among these variables, artificial
missingness (0\%, 15\%, or 30\% additional missing values) was induced
based on patient age, with younger and older patients more likely to
have missing data compared to patients who were between 40 and 65 years
of age. Prior to imputation, 50 predictor variables were selected using
a boosting model that quantified variable importance as the fractional
contribution of each predictor to the model based on the total gain
attributed to using the predictor while growing decision trees.
Imputation was conducted in the training and testing sets, separately,
for each imputation strategy. Although some imputation strategies
(\textit{e.g., }KNN and random forests) can impute data in the testing
set using models fitted to the training set, others
(\textit{e.g., }Bayesian regression, PMM, and hot deck) cannot.
Therefore, to ensure fair comparisons in our experiment, each imputation
procedure imputed data in the training set using models fitted to the
training set and then imputed data in the testing set using models
fitted to the testing set. After imputation, Cox PH and boosting models
were applied to each imputed dataset, separately. Last, model
predictions for death and transplant were computed at 6 months following
MCS surgery.

\paragraph{Bayesian analysis of model performance}

To determine whether any of the imputation methods described above had
improved upon imputation to the mean, we applied Bayesian hierarchical
models to analyze differences in scaled Brier score
\cite{benavoli2017time}. This strategy provides a flexible framework to
conduct hypothesis testing and also accounts for correlation occurring
within each replicate of Monte-Carlo cross validation. Specifically,
within each replicate of Monte-Carlo cross validation, the performance
of different modeling algorithms are correlated because they are trained
and tested using the same data.

@@

Participant characteristics were tabulated for the overall population
and stratified by event status. Continuous and categorical variables
were summarized as mean (standard deviation) and percent, respectively.
The number and proportion of missing values were also tabulated for the
overall population and stratified by event status. Missing data patterns
were visualized for the overall population using an UpSet plot
\cite{lex2014upset}. All of the proceeding analyses were conducted using
resampling results from Monte-Carlo cross validation. Imputation
accuracy was aggregated for all numeric and nominal variables to create
two overall scores for each imputed dataset. For imputation methods that
created multiple datasets, scores were aggregated over each dataset to
provide a single summary score. The distribution
(\textit{i.e., }25\textsuperscript{th}, 50\textsuperscript{th}, and
75\textsuperscript{th} percentile) of the scaled Brier score was
estimated for each modeling algorithm.

We split our results from Monte-carlo cross validation into four
datasets based on outcome (mortality or transplant) and modeling
procedure (Cox PH or boosting). For each dataset, we fit one
hierarchical Bayesian model where the dependent variable was scaled
Brier score and independent variables included the imputation strategy
applied and the amount of artificially missing data (0\%, 15\%, or 30\%)
induced before imputation. With each model, we estimated the difference
in scaled Brier score of downstream models when random forests, Bayesian
regression, PMM, hot deck, MIA, or KNN imputation were applied to impute
missing values instead of imputation to the mean.

@@

Our source code is available on GitHub (see
\url{https://github.com/bcjaeger/INTERMACS-missing-data}).

@@

\paragraph{Missing data}

Many predictor variables exhibited similar proportions of missing values
in different outcome groups (Table \ref{tbl_missingness}). However, the
number (percent) of missing values for surgery time was an exception,
with 6,125 (41.6\%) in the overall population, 2,246 (53.4\%) among
patients who died, and 1,713 (27.2\%) among patients who were censored.
Additionally, missing values for surgery time were frequently
accompanied by missing values for CV pressure (Figure \ref{fig:upset}).

@@

Single imputation strategies were more accurate than their counterparts
using multiple imputation in 8 out of 10 comparisons of nominal scores
and 10 out of 10 comparisons of numeric scores (Table
\ref{tbl_impute_accuracy}). When an additional 15\% of data were
missing, single imputation KNN obtained the highest nominal accuracy
(score: +0.02, 95\% CI -0.05, 0.15) and single imputation with random
forests obtained the highest numeric accuracy (score: +0.05, 95\% CI
0.03, 0.08). When an additional 30\% of data were missing, imputation to
the mean obtained the highest numeric and nominal scores.

\hypertarget{scaled-brier-score}{%
\subsection{Scaled Brier score}\label{scaled-brier-score}}

\paragraph{Mortality risk prediction}

When no additional data were amputed and imputation to the mean was
applied before fitting risk prediction models, the median
(25\textsuperscript{th}, 75\textsuperscript{th} percentile) scaled Brier
score was 6.09 (5.71, 6.54) for Cox PH and 7.27 (6.83, 7.77) for
boosting models (Table \ref{tbl_md_strat_dead_ipa}; top panel). Multiple
imputation strategies universally obtained higher scaled Brier scores
versus their single imputation counterparts. Multiple imputation using
random forests provided the highest scaled Brier score compared to other
strategies, leading to a median (25\textsuperscript{th},
75\textsuperscript{th} percentile) increase in the scaled Brier score of
0.34 (0.12, 0.54) for Cox PH and 0.34 (0.14, 0.58) for boosting models.
These performance increments improved when an aditional 15\% and 30\% of
data were amputed (Table \ref{tbl_md_strat_dead_ipa}; middle and bottom
panel).

\paragraph{Transplant risk prediction}

When no additional data were amputed and imputation to the mean was
applied before fitting risk prediction models, the median
(25\textsuperscript{th}, 75\textsuperscript{th} percentile) scaled Brier
score was 9.35 (8.80, 9.71) for Cox PH and 9.00 (8.55, 9.41) for
boosting models (Table \ref{tbl_md_strat_txpl_ipa}; top panel). For Cox
PH models, imputation to the mean provided the lowest scaled Brier
score, and random forest imputation led to a median
(25\textsuperscript{th}, 75\textsuperscript{th} percentile) increase of
0.13 (0.03, 0.29) versus imputation to the mean. For boosting models,
surprisingly, imputation to the mean provided a higher scaled Brier
score than all imputation methods except for multiple imputation using
Bayesian regression. When an additional 15\% and 30\% of data were
amputed, the performance increments corresponding to the use of multiple
imputation increased for both Cox PH and boosting models (Table
\ref{tbl_md_strat_txpl_ipa}; middle and bottom panel).

\hypertarget{discrimination-and-calibration}{%
\subsection{Discrimination and
calibration}\label{discrimination-and-calibration}}

\paragraph{Mortality risk prediction}

Regardless of how much additional data were amputed, boosting models
obtained higher median C-indices than PH models for prediction of
6-month mortality risk (Table \ref{tbl_md_strat_dead_auc}). In addition,
all models that used multiple imputation consistently obtained higher
median C-indices compared to their counterparts using single imputation.
When 0 and 15\% additional data were amputed, median calibration error
was lower for PH models compared to boosting models, but boosting models
obtained lower median calibration error when 30\% additional data were
amputed (Table \ref{tbl_md_strat_dead_cal_error}). Almost all imputation
strategies provided lower median calibration error compared with
imputation to the mean.

\paragraph{Transplant risk prediction}

For all amounts of additional data amputed, PH models obtained higher
median C-indices than boosting models for prediction of 6-month
transplant risk (Table \ref{tbl_md_strat_txpl_auc}). Similar to
mortality risk prediction, multiple imputation strategies generally
provided higher C-indices than their counterparts using single
imputation strategy. For boosting models, MIA provided higher C-indices
compared to all other single imputation strategies and had similar or
superior performance compared to several multiple imputation strategies.
Differences in calibration error were minor when no additional missing
data were amputed (Table \ref{tbl_md_strat_txpl_cal_error}). When an
additional 30\% of data were amputed, boosting models using MIA obtained
lower calibration error than any other strategy.

\paragraph{Bayesian analysis of model performance}

Adjusting for the amount of additional missing data amputed and the
outcome variable, the posterior probability that an imputation strategy
would improve the scaled Brier score of a downstream model relative to
imputation to the mean was maximized by using multiple imputation with
random forests (probability of improvement: 0.998; Figure
\ref{fig:fig_md_strat_infer_ipa}). Similarly, multiple imputation using
random forests was estimated to have the highest posterior probability
of improving the C-index in comparison to using imputation to the mean
(probability of improvement: 0.962; Figure
\ref{fig:fig_md_strat_infer_auc}). However, the estimated posterior
probability of a reduction in calibration was 1 when using either single
or multiple imputation with random forests compared with imputation to
the mean (Figure \ref{fig:fig_md_strat_infer_cal_error}). Although
imputation using random forest was estimated to be the best overall
option, there was moderate to strong evidence that MIA and each multiple
imputation strategy we applied would improve prognostic accuracy of
downstream models compared with imputation to the mean.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

In this article, we leveraged data from the INTERMACS registry to
evaluate how the use of different imputation strategies prior to fitting
a risk prediction model would impact the external prognostic accuracy of
the model. External prognostic accuracy was measured at 6 months after
receiving MCS to focus on short term risk prediction, and the primary
measure of accuracy was the scaled Brier score. We evaluated the
performance of 12 imputation strategies in a broad range of settings by
varying the type of modeling algorithm, the outcome variable, and the
amount of additional missing data amputed prior to performing
imputation. Our resampling experiment indicated that conducting multiple
imputation has a high likelihood of increasing the downstream scaled
Brier score and C-index of risk prediction models compared with
imputation to the mean. Additionally, multiple imputation with random
forests emerged as the imputation strategy that maximized the
probability of developing a more prognostic model compared with
imputation to the mean.

In previous studies involving the INTERMACS data registry, imputation to
the mean has been applied prior to developing a mortality risk
prediction model
\cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society}.
The current analysis has shown that the prognostic accuracy of these
models can be increased by implementing a different strategy to impute
missing values, particularly if multiple imputation is applied. However,
few studies have investigated best practices on the use of multiple
imputation for prediction models, and few software packages allow users
to implement multiple imputation procedures that adhere to best
practices for prediction model development (\textit{e.g., }using only
the training data to impute missing values). Further development of open
source software that allows investigators to safely and succinctly
implement multiple imputation for prediction models (\textit{e.g., }the
\texttt{miceRanger} package) may increase the presence of multiple
imputation in published risk prediction models.

@@

\begin{tabular}{lccccc}
\toprule
Characteristic & Overall & Censored & Cessation of support & Dead & Transplant\\
\midrule
No. of patients & 14,738 & 6,308 & 310 & 4,204 & 3,916\\
months\_post\_implant & 18 (16) & 25 (18) & 16 (11) & 13 (14) & 13 (10)\\
Age, years & 57 (13) & 57 (13) & 48 (14) & 61 (12) & 53 (12)\\
Sex &  &  &  &  & \\
\hspace{1em}Female & 21 & 22 & 41 & 22 & 19\\
\addlinespace
\hspace{1em}Male & 79 & 78 & 59 & 78 & 81\\
Race &  &  &  &  & \\
\hspace{1em}White & 66 & 63 & 64 & 71 & 67\\
\hspace{1em}Black & 25 & 28 & 27 & 21 & 23\\
\hspace{1em}Other & 9.2 & 9.6 & 9.4 & 7.4 & 10\\
\addlinespace
Body mass index & 29 (7) & 29 (7) & 29 (8) & 29 (7) & 28 (6)\\
Device strategy &  &  &  &  & \\
\hspace{1em}Other & 0.6 & 0.6 & 2.6 & 0.5 & 0.6\\
\hspace{1em}Bridge to transplant & 52 & 43 & 51 & 40 & 81\\
\hspace{1em}Destination therapy & 47 & 56 & 46 & 60 & 18\\
\addlinespace
LVEDD & 6.83 (1.10) & 6.85 (1.07) & 6.49 (0.93) & 6.69 (1.12) & 6.95 (1.12)\\
Urinary albumin, g/dl & 3.41 (0.64) & 3.42 (0.61) & 3.41 (0.64) & 3.33 (0.66) & 3.49 (0.65)\\
Urinary creatinine, mg/dl & 1.39 (0.68) & 1.36 (0.62) & 1.17 (0.53) & 1.49 (0.77) & 1.35 (0.67)\\
CV pressure & 10.8 (6.2) & 10.4 (6.1) & 9.9 (6.0) & 11.6 (6.5) & 10.7 (6.0)\\
Periphal edema & 35 & 33 & 37 & 44 & 31\\
\addlinespace
BUN, mg/dl & 29 (18) & 28 (17) & 22 (13) & 32 (19) & 28 (18)\\
Bilirubin levels, mg/dl & 1.37 (1.84) & 1.25 (1.14) & 1.28 (1.95) & 1.52 (2.46) & 1.41 (1.97)\\
Device type &  &  &  &  & \\
\hspace{1em}Bi-ventricular assistance device & 3.5 & 2.1 & 0.3 & 5.8 & 3.4\\
\hspace{1em}Left-ventricular assistance device & 97 & 98 & 100 & 94 & 97\\
\addlinespace
Surgery time, minutes & 292 (113) & 289 (110) & 272 (96) & 320 (128) & 272 (98)\\
CPB time, minutes & 95 (49) & 93 (45) & 88 (38) & 103 (56) & 90 (44)\\
\bottomrule
\multicolumn{6}{l}{\rule{0pt}{1em}\textit{Note: }}\\
\multicolumn{6}{l}{\rule{0pt}{1em}Table values are mean (standard deviation) or \%}\\
\end{tabular}

@@

\begin{table} 
 \caption{Number (percent) of missing values for a selection of predictor variables in the overall population and in subgroups based on event status.} 
\label{tbl_missingness}

\begin{tabular}{lccccc}
\toprule
 & Overall & Dead & Transplant & Censored & Cessation of support \\ 
\midrule
CV pressure & 7,415 (50.3\%) & 2,156 (51.3\%) & 2,006 (51.2\%) & 3,108 (49.3\%) & 145 (46.8\%) \\ 
Surgery time, minutes & 6,125 (41.6\%) & 2,246 (53.4\%) & 2,007 (51.3\%) & 1,713 (27.2\%) & 159 (51.3\%) \\ 
Periphal edema & 5,093 (34.6\%) & 1,543 (36.7\%) & 1,470 (37.5\%) & 1,952 (30.9\%) & 128 (41.3\%) \\ 
LVEDD & 3,155 (21.4\%) & 1,000 (23.8\%) & 866 (22.1\%) & 1,233 (19.5\%) & 56 (18.1\%) \\ 
Urinary albumin, g/dl & 1,014 (6.88\%) & 326 (7.75\%) & 309 (7.89\%) & 359 (5.69\%) & 20 (6.45\%) \\ 
CPB time, minutes & 1,010 (6.85\%) & 287 (6.83\%) & 274 (7.00\%) & 421 (6.67\%) & 28 (9.03\%) \\ 
Bilirubin levels, mg/dl & 850 (5.77\%) & 291 (6.92\%) & 237 (6.05\%) & 304 (4.82\%) & 18 (5.81\%) \\ 
Body mass index & 77 (0.52\%) & 26 (0.62\%) & 19 (0.49\%) & 31 (0.49\%) & 1 (0.32\%) \\ 
BUN, mg/dl & 50 (0.34\%) & 10 (0.24\%) & 8 (0.20\%) & 32 (0.51\%) & 0 (0.00\%) \\ 
Sex & 24 (0.16\%) & 9 (0.21\%) & 4 (0.10\%) & 11 (0.17\%) & 0 (0.00\%) \\ 
Urinary creatinine, mg/dl & 20 (0.14\%) & 8 (0.19\%) & 4 (0.10\%) & 8 (0.13\%) & 0 (0.00\%) \\ 
Age, years & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
Race & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
Device strategy & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
Device type & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) & 0 (0.00\%) \\ 
\bottomrule
\end{tabular} 
 \end{table}

@@

\begin{tabular}{lcccc}
\toprule
& \multicolumn{2}{c}{Nominal variables} & \multicolumn{2}{c}{Numeric variables} \\ 
 \cmidrule(lr){2-3}\cmidrule(lr){4-5}
 & Single imputation & Multiple imputation & Single imputation & Multiple imputation \\ 
\midrule
\multicolumn{1}{l}{+15\% additional missing data} \\ 
\midrule
Imputation to the mean & 0 (reference) & --- & 0 (reference) & --- \\ 
Hot deck & -0.24 (-0.34, -0.16) & -0.25 (-0.37, -0.18) & -0.35 (-0.53, -0.29) & -0.42 (-0.54, -0.36) \\ 
K-nearest-neighbors & +0.02 (-0.05, 0.15) & -0.01 (-0.09, 0.11) & +0.01 (-0.01, 0.03) & -0.02 (-0.05, 0.00) \\ 
Predictive mean matching & -0.15 (-0.34, 0.00) & -0.15 (-0.35, 0.00) & -0.25 (-0.40, -0.20) & -0.26 (-0.35, -0.21) \\ 
Random forest & -0.15 (-0.26, -0.03) & -0.15 (-0.26, -0.03) & +0.05 (0.03, 0.08) & +0.02 (-0.01, 0.04) \\ 
Bayesian regression & -0.14 (-0.26, 0.01) & -0.14 (-0.27, 0.01) & -0.30 (-0.41, -0.23) & -0.30 (-0.41, -0.23) \\ 
\midrule
\multicolumn{1}{l}{+30\% additional missing data} \\ 
\midrule
Imputation to the mean & 0 (reference) & --- & 0 (reference) & --- \\ 
Hot deck & -0.25 (-0.34, -0.18) & -0.26 (-0.36, -0.18) & -0.34 (-0.46, -0.30) & -0.40 (-0.53, -0.36) \\ 
K-nearest-neighbors & -0.04 (-0.09, 0.03) & -0.07 (-0.13, 0.00) & -0.02 (-0.03, -0.01) & -0.05 (-0.07, -0.04) \\ 
Predictive mean matching & -0.18 (-0.38, -0.04) & -0.18 (-0.37, -0.04) & -0.29 (-0.46, -0.24) & -0.29 (-0.40, -0.25) \\ 
Random forest & -0.18 (-0.28, -0.07) & -0.18 (-0.28, -0.08) & -0.01 (-0.03, 0.01) & -0.03 (-0.06, -0.01) \\ 
Bayesian regression & -0.17 (-0.29, -0.04) & -0.17 (-0.29, -0.05) & -0.31 (-0.43, -0.27) & -0.31 (-0.43, -0.27) \\ 
\bottomrule
\end{tabular} 
 \end{table}

\begin{table} 
 \caption{Median (25th, 75th percentile) change in scaled Brier score when different imputation strategies are applied to training and testing sets instead of imputation to the mean prior to developing a risk prediction model for mortality. Table values show the scaled Brier score for imputation to the mean. For other imputation strategies, table values show the change in scaled Brier score relative to the scaled Brier score when imputation to the mean was applied. All table values are multiplied by 100 for ease of interpretability} 
\label{tbl_md_strat_dead_ipa}
